---
title: "DasyPop methods"
author: "Quentin D. Read"
date: "10/25/2021"
output: html_document
---

# Methods

## Overview

To create a dasymetric population map of each county and county equivalent in the contiguous United States, we took the following approach.
First, we obtained the five-year population estimates from the 2016 American Community Survey (ACS) for each Census block group in each county.
Using imperviousness data from the National Land Cover Database (NLCD), we masked out all 30-m pixels with no impervious surface area, pixels
containing roads, and pixels in Census blocks with zero population. We apportioned the population of each block group among the remaining 
pixels based on the total amount of impervious surface area contained within each pixel. This resulted in a 30-m resolution dasymetric population map.

## Data sources

Our dasymetric population estimation method uses data provided by the U.S. Census Bureau and the Multiresolution Land Characteristics (MRLC) Consortium.
The Census data products we used are the 5-year population estimates at the U.S. Census block group level derived from the 2016 ACS 
and population counts at the U.S. Census block level derived from the 2010 decennial Census. We obtained all Census data from the U.S. Census Bureau API. 
We also obtained the geographical boundaries of blocks and block groups from the Census API.

The MRLC data products we used are the NLCD Percent Developed Imperviousness product for 2016 for the contiguous United
States, and the NLCD Impervious Surface Descriptor product for 2016 for the contiguous United States. See Table.

### Table of data sources

| Data product | Provider | Data year | Resolution | Coverage | Native CRS | Source URL | Date accessed |
| ------------ | -------- | --------- | ---------- | -------- | ---------- | ---------- | ------------- |
| Block group 5-year population estimates, American Community Survey | U.S. Census Bureau | 2016 | &mdash; | contiguous U.S. | &mdash; | https://www.census.gov/programs-surveys/acs | 14 July 2021 | 
| Block population counts, decennial Census | U.S. Census Bureau | 2010 | &mdash; | contiguous U.S. | &mdash; | https://www.census.gov/programs-surveys/decennial-census/decade.2010.html | 14 July 2021 | 
| Census block and block group geographical boundaries | U.S. Census Bureau | 2016 | &mdash; | contiguous U.S. | unprojected latitude-longitude | https://www.census.gov/geographies/mapping-files/time-series/geo/tiger-line-file.2016.html | 14 July 2021 | 
| Impervious surface area | National Land Cover Database | 2016 | 30 m | contiguous U.S. | Albers equal-area | https://www.mrlc.gov/data/nlcd-2016-percent-developed-imperviousness-conus | 13 November 2020 |
| Impervious surface descriptor | National Land Cover Database | 2016 | 30 m | contiguous U.S. | Albers equal-area | https://www.mrlc.gov/data/nlcd-2019-developed-imperviousness-descriptor-conus | 12 July 2021 |

We performed the following steps for each county and county equivalent in the contiguous United States.

## Spatial processing

After obtaining the Census Bureau data, we projected block group and block geographies to the Albers equal-area projection used by NLCD.
As a data cleaning step, we removed polygons with zero area from the block group geographies.
Next, we clipped the NLCD impervious surface percentage and descriptor rasters to the county area (the spatial union of all block groups in the county).

## Masking by impervious surface

We set all impervious surface pixels with a value less than or equal to 0.01 to 0 (less than 1% of the pixel's surface area is impervious).
All pixels in census blocks with zero population in the 2010 decennial census were set to 0 regardless of what impervious surfaces they contained.
Using the impervious surface descriptor, we set all pixels classified as primary, secondary, and tertiary urban roads to 0.

## Dasymetric population density estimation

We rasterized the block group population polygons to match the resolution and extent of the raster containing the remaining impervious surface area pixels 
(pixels remaining after the masking procedure described above).
For each of these impervious surface area pixels, we multiplied the total block group population times the fraction of the block group's 
unmasked impervious surface area contained within that pixel. 
For example if an impervious surface area pixel had a value of 0.5 (50% of the 900 m<sup>2</sup> pixel area is impervious), 
the pixel-level sum of impervious surface area after masking for the block group was 400 (summing the proportions pixel-wise, 
meaning there was a total of 400 * 900 or 360,000 m<sup>2</sup> impervious surface area within the block group), 
and the 2016 population of the entire block group was 4000, 
the dasymetric population density estimate for that pixel would be 4000 * 0.5 / 400 = 5 people per 900 m<sup>2</sup> pixel.

## Correction for anomalous counties

Because we used Census Bureau data products from 2010 and 2016, there are two inconsistencies. 
Between 2010 and 2016, Shannon County, South Dakota was renamed Oglala Lakota County, resulting in a new FIPS code assignment, 
and the independent city of Bedford, Virginia merged with its surrounding county Bedford County. 
We corrected these anomalies manually. 
In the case of Oglala Lakota County, we used the 2016 block group population estimates for Oglala Lakota County and the 2010 block population counts for Shannon County. 
In the case of Bedford County, we used the 2016 block group population estimates for Bedford County (which includes Bedford City) 
and we merged the 2010 block population counts for Bedford City and Bedford County.

## Computational infrastructure

We ran the code in R 4.0.3 and GDAL 2.2.2 on a Slurm cluster running the Linux-based operating system Ubuntu 16.04. 
The code ran in approximately 24 hours across 7 cluster nodes, each with 8 processor cores.

# Code

The following code carries out the steps described above. 

## Initial setup: file tree structure

The following code will set up a directory structure in the local working directory in which to download the external data and write the output. Please modify as necessary.

```{r, eval = FALSE}
dir.create('input_data/NLCD')
dir.create('output_tifs')
dir.create('temp_files')
```

## Pre-processing: download NLCD rasters and create virtual raster

The following Bash code downloads and unzips the two NLCD rasters (impervious surface and impervious surface descriptor).

```{bash, eval = FALSE}
wget https://s3-us-west-2.amazonaws.com/mrlc/nlcd_2016_impervious_l48_20210604.zip -P input_data/NLCD
wget https://s3-us-west-2.amazonaws.com/mrlc/nlcd_2016_impervious_descriptor_l48_20210604.zip -P input_data/NLCD
unzip input_data/NLCD/nlcd_2016_impervious_l48_20210604.zip
unzip input_data/NLCD/nlcd_2016_impervious_descriptor_l48_20210604.zip
```

The following code creates virtual rasters (`.vrt`) from the `.img` raster file. This allows parallel reads to be done on the NLCD raster so that multiple counties can be processed at the same time.

```{r, eval = FALSE}
library(gdalUtils)

direc <- 'input_data/NLCD'
imp_raster_file <- 'nlcd_2016_impervious_l48_20210604.img'
imp_desc_raster_file <- 'nlcd_2016_impervious_descriptor_l48_20210604.img'

# Run gdalbuildvrt with all defaults to just create a VRT that points to the IMG without any modification
gdalbuildvrt(gdalfile = file.path(direc, imp_raster_file), output.vrt = file.path(direc, 'NLCD_2016_impervious.vrt'))
gdalbuildvrt(gdalfile = file.path(direc, imp_desc_raster_file), output.vrt = file.path(direc, 'NLCD_2016_impervious_descriptor.vrt'))
```

## Function to create dasymetric population map for a single county

The following function takes four inputs: the state FIPS code as a two-character string `stid`, the county FIPS code as a three-character string `ctyid`, the file path to the impervious surface raster `imp_raster_file`, and the path to the impervious surface descriptor `imp_desc_raster_file`.
It does the following steps for a single county (described more fully above):

- Obtains 2016 ACS block group populations and 2010 decennial Census block populations using Census Bureau API
- Removes invalid geometries and projects block group geographies to Albers equal-area
- Masks 2016 NLCD impervious percentage and impervious descriptor rasters to the county area
- Sets pixels to zero population that are <1% impervious surface, in blocks with zero population in 2010, or identified as roads
- Sums remaining impervious surface area for each block group
- For each remaining impervious surface area pixel, generates dasymetric population estimate by multiplying block group population by fraction of block group impervious area contained within that pixel
- Writes the result to a GeoTIFF file

*Note*: This requires a valid U.S. Census Bureau API key to be saved in the current working directory in a file named `censusapikey.txt`.
Sign up for a key using [the Census API key signup form](https://api.census.gov/data/key_signup.html).

```{r, eval = FALSE}
Get_Dasy_Data <- function(stid, ctyid, imp_raster_file, imp_desc_raster_file) {
  
  # Albers equal-area projection
  aea <- '+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=WGS84 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs'
  
  # Read Census Bureau API key and get 2016 ACS 5-year block group population estimates
  census_api_key(readLines('censusapikey.txt')) 

  pop <- get_acs(geography = "block group", variables = "B01003_001", 
                 year = 2016, state= stid, county = ctyid, 
                 geometry = TRUE)   
  
  # Data QC: remove empty geometries from pop
  pop <- pop[!is.na(st_dimension(pop)), ]
  
  # Project population to Albers equal-area
  pop.projected <- st_transform(pop, crs = aea)
  
  # Use gdalwarp to extract the county area, from the NLCD impervious percentage raster, already in Albers projection
  temp_polygon_filename <- as.character(glue("temp_files/county-{stid}-{ctyid}.gpkg"))
  temp_nlcdraster_filename <- as.character(glue("temp_files/countynlcd-{stid}-{ctyid}.tif"))
  st_write(st_union(pop.projected), dsn = temp_polygon_filename, driver = 'GPKG')
  gdalwarp(srcfile = imp_raster_file, dstfile = temp_nlcdraster_filename, cutline = temp_polygon_filename, crop_to_cutline = TRUE, tr = c(30, 30), dstnodata = "None")
  lu <- raster(temp_nlcdraster_filename)
  
  # Get 2010 decennial block-level population counts
  # Filter for only the blocks with 0 population, and project to Albers equal-area
  zero.pop <- get_decennial(geography = "block", variables = "P001001", 
                            year = 2010, state = stid, county = ctyid, 
                            geometry = TRUE) %>% 
    filter(value == 0) %>% 
    st_transform(crs = aea)
  
  # Mask NLCD impervious raster to county boundaries
  lu <- mask(lu, as(pop.projected, "Spatial"))
  # Set pixels with impervious percentage <= 1% to 0
  lu[lu <= 1] <- 0
  
  # Scale impervious percentages between 0 and 1
  lu.ratio <- lu/100
  
  # Set all pixels in zero-population blocks to 0
  lu.ratio.zp <- mask(lu.ratio, as(zero.pop, "Spatial"), inverse = TRUE, updatevalue = 0)
  
  # Load impervious surface descriptor dataset, mask all pixels outside the county to NA
  imp.surf.desc <- raster(imp_desc_raster_file)
  imp.surf.crop <- crop(imp.surf.desc, as(pop.projected, "Spatial")) 
  imp.surf.mask <- mask(imp.surf.crop, as(pop.projected, "Spatial")) 

  # Mask out primary, secondary, and urban tertiary roads
  # Reclassify: keep classes 1-6 (non-road) and drop 7-14 (road)
  reclass.table <- matrix(c(1,6,1,7,14,NA), ncol = 3, byrow = TRUE) 
  
  # Reclassify descriptor file and reproject it.
  imp.roads <- reclassify(imp.surf.mask, reclass.table, right = NA)
  imp.roads.p <- projectRaster(imp.roads, lu.ratio.zp, method = 'ngb') 
  # Set all road pixels to 0 (all non-NA values in imp.roads.p)
  RISA <- overlay(lu.ratio.zp, imp.roads.p, fun = function(x, y) {
    x[!is.na(y[])] <- 0
    return(x)
  })
  
  # Get the block group-level sum of the remaining impervious surface pixels
  RISA.sum <- raster::extract(RISA, as(pop.projected,"Spatial"), fun=sum, na.rm=TRUE, df=TRUE)
  
  # Rasterize the block group population estimates and impervious surface pixel sums
  pop.df <- cbind(pop.projected, RISA.sum$layer)
  bg.sum.pop <- fasterize(pop.projected, RISA, field = "estimate")
  bg.sum.RISA <- fasterize(pop.df, RISA, field = "RISA.sum.layer")
  
  # Generate density (people/30 m pixel) and write to file.
  dasy.pop <- (bg.sum.pop/bg.sum.RISA) * RISA
  
  filename <- glue("output_tifs/neon-dasy-{stid}-{ctyid}.tif")
  
  writeRaster(dasy.pop, filename, overwrite = TRUE, NAflag = -9999) 
  
  message(glue("saved raster with stid {stid} and ctyid {ctyid}. On to the next one!"))
  
}
```

## Setup

The following code block loads the necessary R packages. The object `fips_codes` is a data frame that loads with the `tidycensus` package. 
It contains the two-digit state FIPS code and the three-digit county FIPS code for each county in the United States.
We removed all counties not in the 48 contiguous United States and District of Columbia. In addition, we removed the two anomalous counties described above.

```{r, eval = FALSE}
# Load packages
library(tidycensus)
library(tidyverse)
library(raster)
library(sf)
library(glue)
library(gdalUtils)
library(rslurm)
library(fasterize)

imp_raster_file <- 'input_data/NLCD/nlcd_2016_impervious_l48_20210604.img'
imp_desc_raster_file <- 'input_data/NLCD/nlcd_2016_impervious_descriptor_l48_20210604.img'

# Get the fips codes for all counties
fipscodes = data_frame(fips_codes)

fipscodes = fipscodes %>% 
  rename(
    stid = state_code, 
    ctyid = county_code
  ) %>% dplyr::select(stid, ctyid)

# Get rid of the ones that are not in the 48 continental states and DC
not48 <- c('02','15','60','66','69','72','74','78')
fipscodes <- filter(fipscodes, !stid %in% not48)

# Remove Shannon County, South Dakota (46113) from the list because it does not exist in 2016 (replaced by 46102, Oglala Lakota County)
# Also remove Bedford City, Virginia (51515) from the list because it does not exist in 2016 (merged into 51019, Bedford County)
fipscodes <- filter(fipscodes, !(stid == '46' & ctyid %in% c('102', '113')) & !(stid == '51' & ctyid %in% c('019', '515')))

# Split into a list by state.
fips_list <- fipscodes %>% group_by(stid) %>% group_split
```

## Apply function to all counties

THe following code block defines a function to wrap around `Get_Dasy_Data()` and run it for each county in a state in succession. 
This is necessary to avoid parallel read on a single raster file (temporary files created by functions in the `tidycensus` package). 
Then, the function is run in parallel on a Slurm cluster.

```{r, eval = FALSE}
# Function to run all counties in a state
get_dasy_all_counties <- function(fips) {
  walk(fips$ctyid, ~ Get_Dasy_Data(stid = fips$stid[1], ctyid = ., 
                                   imp_raster_file = imp_raster_file,
                                   imp_desc_raster_file = imp_desc_raster_file))
}

sjob <- slurm_map(fips_list, get_dasy_all_counties, jobname = 'DASYallcounties',
                  nodes = 7, cpus_per_node = 8, pkgs = c("tidycensus", "raster", "tidyverse", "sf", "glue", "gdalUtils"),
                  global_objects = "Get_Dasy_Data")

cleanup_files(sjob, wait = TRUE)
# Also delete the temporarily downloaded files
system2('rm', '-r temp_files/*') 
```

## Corrections for two inconsistent counties

The following code block manually fixes the inconsistencies in Oglala Lakota County, South Dakota, and Bedford County, Virginia, as described above. 
Refer to pg. 3 of [this documentation](https://www.ddorn.net/data/FIPS_County_Code_Changes.pdf) to see what happened to those counties since 2010.
For Oglala Lakota, we download data for FIPS code 46113 for 2010, and 46102 for 2016.
For Bedford, we merge data from 51515 and 51019 for 2010, and download data for 51019 for 2016.

```{r, eval = FALSE}
bad_counties <- data.frame(stid = c('46','46','51'),
                           ctyid = c('102', '113', '515'))

census_api_key(readLines('censusapikey.txt'))

# Manual fix for South Dakota ---------------------------------------------

# 2016 data is new code.
stid <- "46"
ctyid <- "102"
pop <- get_acs(geography = "block group", variables = "B01003_001", 
               year = 2016, state= stid, county = ctyid, 
               geometry = TRUE)   

aea <- '+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=WGS84 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs'

# Data QC: remove empty geometries from pop
pop <- pop[!is.na(st_dimension(pop)), ]
pop.projected <- st_transform(pop, crs = aea)

# Still use 46102 for NLCD (2016)
temp_polygon_filename <- as.character(glue("temp_files/county-{stid}-{ctyid}.gpkg"))
temp_nlcdraster_filename <- as.character(glue("temp_files/countynlcd-{stid}-{ctyid}.tif"))
st_write(st_union(pop.projected), dsn = temp_polygon_filename, driver = 'GPKG')
gdalwarp(srcfile = imp_raster_file, dstfile = temp_nlcdraster_filename, cutline = temp_polygon_filename, crop_to_cutline = TRUE, tr = c(30, 30), dstnodata = "None")
lu <- raster(temp_nlcdraster_filename)

# Use old FIPS code for 2010
ctyid <- "113"
# Get 2010 decennial block-level population counts
# Filter for only the blocks with 0 population, and project to Albers equal-area
zero.pop <- get_decennial(geography = "block", variables = "P001001", 
                          year = 2010, state = stid, county = ctyid, 
                          geometry = TRUE) %>% 
  filter(value == 0) %>% 
  st_transform(crs = aea)

# Mask NLCD impervious raster to county boundaries
lu <- mask(lu, as(pop.projected, "Spatial"))
# Set pixels with impervious percentage <= 1% to 0
lu[lu <= 1] <- 0

# Scale impervious percentages between 0 and 1
lu.ratio <- lu/100

# Set all pixels in zero-population blocks to 0
lu.ratio.zp <- mask(lu.ratio, as(zero.pop, "Spatial"), inverse = TRUE, updatevalue = 0)

# Load impervious surface descriptor dataset, mask all pixels outside the county to NA
imp.surf.desc <- raster(imp_desc_raster_file)
imp.surf.crop <- crop(imp.surf.desc, as(pop.projected, "Spatial")) 
imp.surf.mask <- mask(imp.surf.crop, as(pop.projected, "Spatial")) 

# Mask out primary, secondary, and urban tertiary roads
# Reclassify: keep classes 1-6 (non-road) and drop 7-14 (road)
reclass.table <- matrix(c(1,6,1,7,14,NA), ncol = 3, byrow = TRUE) 

# Reclassify descriptor file and reproject it.
imp.roads <- reclassify(imp.surf.mask, reclass.table, right = NA)
imp.roads.p <- projectRaster(imp.roads, lu.ratio.zp, method = 'ngb') 
# Set all road pixels to 0 (all non-NA values in imp.roads.p)
RISA <- overlay(lu.ratio.zp, imp.roads.p, fun = function(x, y) {
  x[!is.na(y[])] <- 0
  return(x)
})

# Get the block group-level sum of the remaining impervious surface pixels
RISA.sum <- raster::extract(RISA, as(pop.projected,"Spatial"), fun=sum, na.rm=TRUE, df=TRUE)

# Rasterize the block group population estimates and impervious surface pixel sums
pop.df <- cbind(pop.projected, RISA.sum$layer)
bg.sum.pop <- fasterize(pop.projected, RISA, field = "estimate")
bg.sum.RISA <- fasterize(pop.df, RISA, field = "RISA.sum.layer")

# Generate density (people/30 m pixel) and write to file.
dasy.pop <- (bg.sum.pop/bg.sum.RISA) * RISA

filename = as.character(glue("output_tifs/neon-dasy-{stid}-102.tif"))

writeRaster(dasy.pop, filename, overwrite = TRUE, NAflag = -9999) # Will overwrite existing file with the same name.


# Manual fix for Virginia -------------------------------------------------

# Only one code used for 2016
stid <- "51"
ctyid <- "019"
pop <- get_acs(geography = "block group", variables = "B01003_001", 
               year = 2016, state= stid, county = ctyid, 
               geometry = TRUE)   

# Data QC: remove empty geometries from pop
pop <- pop[!is.na(st_dimension(pop)), ]
pop.projected <- st_transform(pop, crs = aea)

temp_polygon_filename <- as.character(glue("temp_files/county-{stid}-{ctyid}.gpkg"))
temp_nlcdraster_filename <- as.character(glue("temp_files/countynlcd-{stid}-{ctyid}.tif"))
st_write(st_union(pop.projected), dsn = temp_polygon_filename, driver = 'GPKG')
gdalwarp(srcfile = imp_raster_file, dstfile = temp_nlcdraster_filename, cutline = temp_polygon_filename, crop_to_cutline = TRUE, tr = c(30, 30), dstnodata = "None")
lu <- raster(temp_nlcdraster_filename)

# Download 2010 block-level data, filter for only the blocks with 0 pop
# We need to get both the city and county and merge them.
zero.pop.bedfordcounty <- get_decennial(geography = "block", variables = "P001001", 
                          year = 2010, state = stid, county = ctyid, 
                          geometry = TRUE) %>% filter(value == 0) %>% st_transform(., proj4string(lu))
zero.pop.bedfordcity <- get_decennial(geography = "block", variables = "P001001", 
                                      year = 2010, state = stid, county = "515", 
                                      geometry = TRUE) %>% filter(value == 0) %>% st_transform(., proj4string(lu))

# Plot to check them

ggplot() +
  geom_sf(data = zero.pop.bedfordcounty, aes(fill = value), col = "blue") +
  geom_sf(data = zero.pop.bedfordcity, aes(fill = value), col = "red") 
# Looks good! There is a hole in the middle of the county filled in with the city.

zero.pop <- bind_rows(zero.pop.bedfordcity, zero.pop.bedfordcounty)

# Mask NLCD impervious raster to county boundaries
lu <- mask(lu, as(pop.projected, "Spatial"))
# Set pixels with impervious percentage <= 1% to 0
lu[lu <= 1] <- 0

# Scale impervious percentages between 0 and 1
lu.ratio <- lu/100

# Set all pixels in zero-population blocks to 0
lu.ratio.zp <- mask(lu.ratio, as(zero.pop, "Spatial"), inverse = TRUE, updatevalue = 0)

# Load impervious surface descriptor dataset, mask all pixels outside the county to NA
imp.surf.desc <- raster(imp_desc_raster_file)
imp.surf.crop <- crop(imp.surf.desc, as(pop.projected, "Spatial")) 
imp.surf.mask <- mask(imp.surf.crop, as(pop.projected, "Spatial")) 

# Mask out primary, secondary, and urban tertiary roads
# Reclassify: keep classes 1-6 (non-road) and drop 7-14 (road)
reclass.table <- matrix(c(1,6,1,7,14,NA), ncol = 3, byrow = TRUE) 

# Reclassify descriptor file and reproject it.
imp.roads <- reclassify(imp.surf.mask, reclass.table, right = NA)
imp.roads.p <- projectRaster(imp.roads, lu.ratio.zp, method = 'ngb') 
# Set all road pixels to 0 (all non-NA values in imp.roads.p)
RISA <- overlay(lu.ratio.zp, imp.roads.p, fun = function(x, y) {
  x[!is.na(y[])] <- 0
  return(x)
})

# Get the block group-level sum of the remaining impervious surface pixels
RISA.sum <- raster::extract(RISA, as(pop.projected,"Spatial"), fun=sum, na.rm=TRUE, df=TRUE)

# Rasterize the block group population estimates and impervious surface pixel sums
pop.df <- cbind(pop.projected, RISA.sum$layer)
bg.sum.pop <- fasterize(pop.projected, RISA, field = "estimate")
bg.sum.RISA <- fasterize(pop.df, RISA, field = "RISA.sum.layer")

# Generate density (people/30 m pixel) and write to file.
dasy.pop <- (bg.sum.pop/bg.sum.RISA) * RISA

filename = as.character(glue("output_tifs/neon-dasy-{stid}-{ctyid}.tif"))

writeRaster(dasy.pop, filename, overwrite = TRUE, NAflag = -9999) # Will overwrite existing file with the same name.
```

## Mosaic rasters

The following code block is a set of Bash shell commands to mosaic the county-level GeoTIFF together as a virtual raster with `gdalbuildvrt`. Next, `gdal_translate` converts the virtual raster to a single large GeoTIFF.

```{bash, eval = FALSE}
cd output_tifs
gdalbuildvrt dasy_conus_2021-08-26.vrt *.tif
gdal_translate -co COMPRESS=LZW -co BIGTIFF=YES -of GTiff dasy_conus_2021-08-26.vrt dasy_conus_2021-08-26.tif
```

